\documentclass[journal]{IEEEtran}

% IEEE TNSM submission draft (>=10pp, transaction-grade).
% Author: Do Phuc Hao (CAIRA, Da Nang Architecture University), sole + corresponding.
% Build: latexmk -pdf main.tex
% All numbers are aggregated from results/*.csv via paper/make_figs_data.py.

\usepackage{amsmath,amssymb,mathtools}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}
\usepgfplotslibrary{statistics}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,fit,backgrounds,calc,shapes.geometric}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{url}
\usepackage[hidelinks]{hyperref}

\tikzset{
  blk/.style={rectangle, rounded corners=2pt, draw=black, fill=blue!8, text=black,
              align=center, minimum height=8mm, inner sep=3pt, font=\small},
  blkb/.style={rectangle, rounded corners=2pt, draw=black, fill=orange!14, text=black,
               align=center, minimum height=8mm, inner sep=3pt, font=\small},
  sat/.style={circle, draw=black, fill=blue!18, text=black, minimum size=3.4mm,
              inner sep=0pt},
  flowarr/.style={-{Latex[length=2mm]}, thick},
}

\newcommand{\cost}{\mathrm{cost}}
\newcommand{\prop}{\mathrm{prop}}
\newcommand{\cpc}{\mathrm{cap}}
\newcommand{\softplus}{\mathrm{softplus}}
\newcommand{\TTT}{\mathrm{TTT}}
\newcommand{\AoN}{\textsc{AoN}}

\begin{document}

\title{Learning to Route Under Load: An Inductive Graph Network that Amortizes
Congestion-Aware Traffic Engineering for LEO Mega-Constellations}

\author{Do~Phuc~Hao% <-this % stops a space
\thanks{D. P. Hao is with the Center for Artificial Intelligence Research and
Application (CAIRA), Da Nang Architecture University, Da Nang, Vietnam
(e-mail: haodp@dau.edu.vn). Corresponding author: D. P. Hao.}}

\markboth{IEEE Transactions on Network and Service Management (submitted)}%
{Hao: Learning to Route Under Load}

\maketitle

\begin{abstract}
Low Earth orbit (LEO) mega-constellations carry traffic over a dense mesh of
inter-satellite links whose topology, although time varying, is known in advance
from orbital ephemeris. Under load, routing on propagation delay alone concentrates
traffic on a few bottleneck links while alternative paths sit idle, wasting most of
the achievable performance: in our study it inflates total travel time by up to a
factor of forty. The congestion-aware optimum, the user equilibrium, removes this
waste but requires the full demand matrix and an iterative solve at every slot, which
does not scale to constellations of more than a thousand satellites with rapidly
changing topology. We formalize routing under load on LEO as the amortization of a
traffic-assignment equilibrium, decompose it into a price-prediction sub-problem, a
decoding sub-problem, an inductive-transfer sub-problem, and a proactive sub-problem,
and solve the first with an inductive graph neural network that predicts the
equilibrium congestion price on each inter-satellite link in a single forward pass
from cheap, observable load features. On a constellation simulator with a standard
Bureau of Public Roads cost model, the network closely approaches the user
equilibrium in distribution (recovering about $92\%$ of the gain over blind routing),
recovers about $71\%$ of that gain when transferred zero shot to a shell twelve times
larger, runs $19$ to $29\times$ faster than the iterative solve with the saving
growing with scale, and, fed predicted demand, sustains low travel time under
drifting hotspots where a reactive controller degrades to $1.6\times$ blind. We
contribute, in addition, a record of three evaluation pitfalls that silently
fabricate or distort such results, with a protocol to avoid them, and we report a
negative result: a quantum-walk propagation operator that motivated this study offers
no benefit once load features are available, and a plain graph convolution is
preferable.
\end{abstract}

\begin{IEEEkeywords}
LEO satellite networks, non-terrestrial networks, routing, traffic engineering,
graph neural networks, congestion control, user equilibrium, inductive learning,
evaluation methodology.
\end{IEEEkeywords}

\section{Introduction}
\IEEEPARstart{L}{ow} Earth orbit (LEO) mega-constellations such as Starlink and
Kuiper interconnect their satellites with laser inter-satellite links (ISLs)
arranged in a regular grid, forming a wide-area packet network that moves with the
constellation. Two features distinguish this network from a terrestrial one. First,
the topology is time varying but \emph{deterministic}: the position of every
satellite, hence every ISL and its propagation delay, is known in advance from the
orbital parameters. Second, the network is large and fast: a single shell contains a
thousand or more satellites, and cross-plane links open and close as satellites
traverse the constellation.

Routing such a network on propagation delay alone is attractive because the shortest
path is cheap and, when lightly loaded, near optimal. Under load it is wasteful: it
funnels many flows onto the same minimum-hop links, creating hotspots, while
geometrically longer but idle paths are ignored. We measure this waste directly
(Section~\ref{sec:headroom}) and find that, once the network is moderately loaded,
congestion-aware routing cuts total travel time by more than half, and by up to a
factor of forty under heavy load.

The congestion-aware optimum is classical. Routing every flow so that none can
unilaterally lower its cost yields the \emph{user equilibrium} (UE), computed by
iterating shortest-path assignment against a load-dependent cost. UE is the right
target but expensive: it needs the entire demand matrix and many shortest-path rounds
per slot. At constellation scale, with topology changing every few seconds,
recomputing UE from scratch each slot is impractical.

This tension, a cheap-but-blind heuristic versus an expensive-but-optimal solve,
motivates our approach: \emph{amortize} the equilibrium. A graph neural network (GNN)
observes cheap features (per-node traffic intensity and the load that one
congestion-blind pass would induce) and predicts, in a single forward pass, the
equilibrium congestion price on every ISL. Routing on the predicted cost recovers
most of the congestion-aware gain at a fraction of the cost of solving UE. Because the
model's parameters do not depend on the number of satellites, it is \emph{inductive}:
trained on a small shell, it applies unchanged to a larger one. And because the demand
pattern drifts predictably (the dense-traffic band rotates under the constellation),
feeding the predicted near future demand lets the router act \emph{proactively},
before congestion forms. Figure~\ref{fig:pipeline} summarizes the pipeline.

To make the stakes concrete, consider the dense-traffic band passing over a populated
longitude. Delay-optimal routing sends the surge of flows along the few minimum-hop ISL
chains that cross that region, driving a handful of links far past capacity while the
cross-plane links a hop away carry almost nothing. With a fourth-power congestion cost, the
travel time on those saturated links explodes, and the network appears to be in overload even
though the same traffic spread across the available ISLs would be comfortably feasible. An
operator that recomputes the equilibrium each slot fixes this but pays a heavy, repeated
optimization; an operator that routes on stale or blind information does not fix it at all.
The method in this paper gives a third option: spend the optimization once, offline, and let a
network predict its outcome thereafter.

A second, methodological thread runs through the paper. Learned traffic engineering is
easy to evaluate incorrectly: a greedy baseline that silently drops hard flows, a
scaling claim that rests on a ratio with a moving denominator, or an equilibrium
reference measured off its operating point can each fabricate or distort a result. We
encountered all three, caught them, and distil a protocol (Section~\ref{sec:pitfalls})
so that the reported numbers are trustworthy and reproducible.

\textbf{Contributions.}
\begin{itemize}
\item We formalize routing under load on LEO as the amortization of a
traffic-assignment equilibrium and \emph{decompose} it into four sub-problems
(Section~\ref{sec:decomp}), each mapped to a measurable claim.
\item \textbf{C1 (amortization).} A one-shot GNN closely approaches the user
equilibrium in distribution, recovering about $92\%$ of the gain over blind routing,
at $19$ to $29\times$ less computation than the iterative solve.
\item \textbf{C2 (inductive scale generalization).} Trained on a $132$-satellite
shell, the same model transferred zero shot to a $1584$-satellite shell ($12\times$
larger) recovers about $71\%$ of the gain and remains the best \emph{deployable}
policy.
\item \textbf{C3 (cost that improves with scale).} We give a complexity analysis and
wall-clock measurements showing the inference saving over UE \emph{grows} with the
constellation, reaching $29\times$ at $1584$ satellites.
\item \textbf{C4 (proactive).} Under drifting hotspots, routing on predicted demand
tracks the clairvoyant optimum, whereas a reactive one-slot-lagged controller degrades
below blind shortest path as the drift grows.
\item \textbf{Evaluation methodology.} We document three pitfalls that silently distort
results in this setting, with fixes, and verify the corrected estimator against the
price-of-anarchy ordering $\TTT(\mathrm{SO})\le\TTT(\mathrm{UE})$.
\item \textbf{Honest negative result.} The study began as a quantum-walk GNN; we report
the kill-gate record and why a plain graph convolution is the right operator here.
\end{itemize}

\section{Related Work}
\textbf{LEO routing and topology.} The deterministic, time-varying structure of LEO
ISL networks has been studied for low-latency routing and topology
design~\cite{handley2018,bhattacherjee2019}. Handley showed that a $+$Grid of ISLs can
beat terrestrial fiber on long paths because light is faster in vacuum, making the ISL
layer worth routing carefully~\cite{handley2018}; Bhattacherjee and Snoeren explored how
the choice of which satellites to interconnect shapes path length and
churn~\cite{bhattacherjee2019}. Packet-level simulators such as Hypatia~\cite{hypatia}
and StarPerf~\cite{starperf} reconstruct the moving topology from constellation
parameters and have driven work on path stability, handover, and delay under realistic
orbital dynamics. This line predominantly routes on delay or hop count and evaluates a
lightly loaded network; the contribution we make is orthogonal and complementary, namely
how to route when the bottleneck is congestion rather than geometry, and how to do so
without a per-slot optimization.

\textbf{GNNs for networking.} Graph convolutions~\cite{gcn}, attention~\cite{gat}, and
inductive sampling~\cite{graphsage} underpin learning on graphs. RouteNet predicts
per-path delay and jitter for network modeling and what-if analysis~\cite{routenet};
reinforcement learning with GNNs has been applied to routing
optimization~\cite{almasan}. These approaches either model performance or learn a
control policy by trial and error. Ours is supervised and predicts a congestion-price
\emph{field} that is decoded by a single shortest-path pass, which keeps inference to
one forward evaluation and makes inductivity across constellation size immediate.

\textbf{Traffic engineering and equilibria.} Load-dependent link cost, the user
equilibrium, and the system optimum are classical in transportation
science~\cite{wardrop,bpr,beckmann}: Wardrop's principles characterize the equilibrium,
the Beckmann transformation casts it as a convex program, and the BPR function is the
standard congestion cost. The gap between selfish and optimal routing is the price of
anarchy, bounded for such cost classes by Roughgarden and Tardos~\cite{poa}. Learning to
route so as to reduce congestion has been explored for terrestrial and software-defined
networks~\cite{valadarsky,xu}, typically with reinforcement learning over a fixed
topology. Two things change in LEO and motivate a different design: the topology is large
and changes every few seconds, so a per-slot equilibrium solve is the bottleneck and
amortizing it is the prize; and the demand is non-stationary but \emph{predictable}, so a
forecast-driven, proactive controller is both possible and valuable. We therefore learn
to predict the equilibrium price field and decode it with a classical router, rather than
learn a topology-specific policy by trial and error.

\textbf{Quantum-walk networks.} Continuous-time quantum walks have been used as a graph
propagation operator to capture long-range structure~\cite{qwnn}. We evaluate one as a
candidate operator and report it as a negative ablation
(Section~\ref{sec:ablation}).

\section{System Model}\label{sec:model}
Table~\ref{tab:notation} lists the notation.

\begin{table}[t]
\centering
\caption{Notation.}
\label{tab:notation}
\begin{tabular}{ll}
\toprule
symbol & meaning \\
\midrule
$G_t=(V,E_t)$ & constellation graph at slot $t$ \\
$\prop_{uv}$ & propagation delay of ISL $(u,v)$ \\
$\cpc$ & ISL capacity \\
$\mathcal{D}=\{(s,d,r)\}$ & demand set (source, destination, rate) \\
$x_{uv}$ & load (aggregate rate) on ISL $(u,v)$ \\
$\cost_{uv}(x)$ & realized link cost under load (BPR) \\
$g^{*}_{uv}$ & user-equilibrium congestion price of $(u,v)$ \\
$\hat g_{uv}$ & price predicted by the GNN \\
$h_v^{(l)}$ & embedding of node $v$ at layer $l$ \\
$\TTT(\pi)$ & total travel time of routing policy $\pi$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Time-varying constellation graph}
A Walker shell of $S$ satellites in $P$ planes is propagated on circular orbits. Each
satellite holds two intra-plane ISLs and up to two cross-plane ISLs, forming a $+$Grid
(Fig.~\ref{fig:topo}). The topology over a horizon is a known sequence
$\{G_t\}_{t=0}^{T-1}$, $G_t=(V,E_t)$, with $|V|=S$. Each ISL $(u,v)\in E_t$ has a
propagation delay $\prop_{uv}=\lVert p_u-p_v\rVert/c$ from the known positions $p$, and
a capacity $\cpc$. Cross-plane links above a polar cutoff latitude are deactivated, as
in operational designs; for a $53^\circ$ shell the cutoff is never reached, so the ISL
graph is near static and the dynamics enter through the demand.

\subsection{Demand, flows, and link load}
A demand set $\mathcal{D}=\{(s_k,d_k,r_k)\}$ requests rate $r_k$ from $s_k$ to $d_k$.
We draw demands from a gravity model: a node's attractiveness is
$w_v\propto\exp(\kappa\cos(\angle v-\phi_t))$, peaked on a longitude band $\phi_t$, and
endpoints are sampled with probability proportional to $w_{s}w_{d}$. The band $\phi_t$
rotates between slots, a proxy for the dense-population and sub-solar region rotating
under the constellation as Earth turns; the per-slot drift $\Delta=\phi_{t+1}-\phi_t$ is
known from the same ephemeris that fixes the topology. This makes the demand
non-stationary (the hotspot moves) yet predictable (its motion is known), which is
exactly the structure the proactive variant exploits. Offered load is controlled by the
number of demands, which we sweep to span the lightly loaded and heavily congested
regimes. A routing
policy maps each demand $k$ to a path $P_k$ in $G_t$. Writing $r_k$ for the rate on
$P_k$, the load on an ISL is
\begin{equation}
x_{uv}=\sum_{k}\ r_k\,\mathbb{1}\!\left[(u,v)\in P_k\right],
\label{eq:load}
\end{equation}
subject to flow conservation at every node $v\notin\{s_k,d_k\}$,
\begin{equation}
\sum_{u:(u,v)\in E_t} f^{k}_{uv}=\sum_{w:(v,w)\in E_t} f^{k}_{vw},
\label{eq:conserv}
\end{equation}
with $f^{k}$ the per-demand flow.

\subsection{Congestion cost and realized travel time}
The realized cost of a link under load follows the Bureau of Public Roads (BPR)
form~\cite{bpr},
\begin{equation}
\cost_{uv}(x)=\prop_{uv}\Big(1+\alpha\,(x_{uv}/\cpc)^{\beta}\Big),
\quad \alpha=0.15,\ \beta=4 ,
\label{eq:bpr}
\end{equation}
which is smooth, strictly increasing, and convex in $x$, and which blows up as a link
approaches and exceeds capacity. The total travel time of a policy $\pi$ inducing loads
$x(\pi)$ is
\begin{equation}
\TTT(\pi)=\sum_{(u,v)}x_{uv}(\pi)\,\cost_{uv}\big(x_{uv}(\pi)\big)
        =\sum_{k} r_k \!\!\sum_{(u,v)\in P_k}\!\!\cost_{uv}\big(x_{uv}(\pi)\big).
\label{eq:ttt}
\end{equation}
The two forms are equal because $\sum_k r_k\sum_{(u,v)\in P_k}(\cdot)=\sum_{(u,v)}x_{uv}(\cdot)$;
the edge form is the one to evaluate at an equilibrium, a point we return to in
Section~\ref{sec:pitfalls}.

\subsection{User equilibrium and system optimum}
Two operating points bound any router. The \emph{user equilibrium} (UE) satisfies
Wardrop's first principle: for every demand, all used paths have equal and minimal
cost. Equivalently, UE is the unique minimizer of the Beckmann
potential~\cite{beckmann}
\begin{equation}
x^{\mathrm{UE}}=\arg\min_{x\in\mathcal{F}}\ \sum_{(u,v)}\int_{0}^{x_{uv}}
\cost_{uv}(w)\,\mathrm{d}w ,
\label{eq:beckmann}
\end{equation}
over the polytope $\mathcal{F}$ of loads induced by feasible flows
\eqref{eq:load}--\eqref{eq:conserv}; with \eqref{eq:bpr} the integral is
$\prop_{uv}\!\big(x+\tfrac{\alpha\,\cpc}{\beta+1}(x/\cpc)^{\beta+1}\big)$. The
\emph{system optimum} (SO) instead minimizes \eqref{eq:ttt} directly,
$x^{\mathrm{SO}}=\arg\min_{x\in\mathcal{F}}\TTT(x)$, equivalently routes on the marginal
cost $\cost_{uv}(x)+x_{uv}\cost'_{uv}(x)$, and lower-bounds UE. Their ratio
$\TTT(\mathrm{UE})/\TTT(\mathrm{SO})$ is the price of anarchy~\cite{poa}. We use UE as
the deployable congestion-aware reference and SO as a tighter ceiling; in our regime
they nearly coincide (Section~\ref{sec:results}).

We compute UE by the method of successive averages (MSA): starting from an
all-or-nothing assignment $y^{0}$ on free-flow cost, iterate
\begin{equation}
x^{k+1}=x^{k}+\tfrac{1}{k+1}\big(y^{k}-x^{k}\big),\quad
y^{k}=\AoN\big(\cost(x^{k})\big),
\label{eq:msa}
\end{equation}
where $\AoN$ routes every demand on the shortest path under the current cost; SO uses
the same iteration on the marginal cost. Each iteration is $|\mathcal{D}|$
shortest paths and convergence needs $T$ iterations.

\subsection{The amortization problem}
Blind shortest path is one $\AoN$ pass and is congestion blind; UE is $T$ passes and is
congestion optimal but does not scale. We seek a parametric map $f_\theta$ from cheap,
observable state $\mathcal{S}_t$ to a routing $\pi_\theta$ that approaches UE in one
forward pass and whose parameters $\theta$ are independent of $S$:
\begin{equation}
\min_{\theta}\ \mathbb{E}_{\mathcal{D},G}\big[\TTT(\pi_\theta)\big]
\quad\text{s.t.}\quad \mathrm{cost}(\pi_\theta)\ll \mathrm{cost}(\mathrm{UE}).
\label{eq:amort}
\end{equation}

\section{Problem Decomposition}\label{sec:decomp}
We decompose \eqref{eq:amort} into four sub-problems (Fig.~\ref{fig:decomp}), each
mapped to a claim.

\noindent\textbf{(SP1) Price prediction.} Approximate the UE congestion price field
\begin{equation}
g^{*}_{uv}=\cost^{\mathrm{UE}}_{uv}/\prop_{uv}-1\ \ge 0,
\qquad \hat g = f_\theta(\mathcal{S}_t),
\label{eq:price}
\end{equation}
so that routing on $\prop(1+\hat g)$ reproduces $x^{\mathrm{UE}}$. Maps to \textbf{C1}.

\noindent\textbf{(SP2) Decoding.} Turn a price field into a routing by one shortest-path
assignment, $\pi=\AoN(\prop(1+\hat g))$. The decode is shared across all demands, so a
single $\hat g$ routes the whole matrix.

\noindent\textbf{(SP3) Inductive transfer.} Require $\theta$ to be size invariant so that
$f_\theta$ trained on $\{G^{\mathrm{small}}\}$ applies to a larger $G^{\mathrm{large}}$
zero shot. Maps to \textbf{C2}. A message-passing $f_\theta$ with shared weights and a
local readout satisfies this by construction (Section~\ref{sec:induct}).

\noindent\textbf{(SP4) Proactivity.} Under drift $\Delta$, predict next-slot demand
$\hat{\mathcal{D}}_{t}$ from $\mathcal{D}_{t-1}$ and the known $\Delta$, and decode on
$f_\theta(\mathcal{S}(\hat{\mathcal{D}}_t))$ so the route pre-empts the moving hotspot.
Maps to \textbf{C4}.

Cost of SP2 plus inference versus the UE solve gives \textbf{C3}
(Section~\ref{sec:complexity}).

\begin{figure*}[t]
\centering
\resizebox{0.96\textwidth}{!}{%
\begin{tikzpicture}[node distance=10mm and 12mm]
\node[blk] (g) {Constellation\\ graph $G_t$};
\node[blk, right=of g] (obs) {Observe\\ demand $\mathcal{D}_t$,\\ blind load $x^{0}$};
\node[blk, right=of obs] (feat) {Node/edge\\ features $\mathcal{S}_t$};
\node[blk, right=of feat] (gnn) {Edge-price GNN\\ $K$ message-passing\\ layers};
\node[blk, right=of gnn] (price) {Edge prices\\ $\hat g_{uv}$};
\node[blk, right=of price] (dec) {Shortest-path\\ decode on\\ $\prop(1+\hat g)$};
\node[blkb, right=of dec] (out) {Routed flows\\ realized $\TTT$};
\foreach \a/\b in {g/obs,obs/feat,feat/gnn,gnn/price,price/dec,dec/out}
  \draw[flowarr] (\a) -- (\b);
\node[blkb, above=8mm of feat] (fc) {Demand forecast\\ $\hat{\mathcal{D}}_t$ (proactive)};
\draw[flowarr, dashed] (fc) -- (feat);
\end{tikzpicture}}
\caption{End-to-end pipeline. Cheap observable state (demand and the load that one
congestion-blind pass would induce) feeds an edge-price GNN that predicts the
equilibrium congestion price per ISL in one forward pass; a single shortest-path decode
then routes the whole demand matrix. The proactive variant (dashed) feeds the predicted
next-slot demand.}
\label{fig:pipeline}
\end{figure*}

\begin{figure}[t]
\centering
\resizebox{0.9\columnwidth}{!}{%
\begin{tikzpicture}[node distance=6mm and 6mm]
\node[blk] (root) {Amortize UE \eqref{eq:amort}};
\node[blk, below left=10mm and -2mm of root] (sp1) {SP1 price\\ prediction (C1)};
\node[blk, below=18mm of root] (sp2) {SP2 decode};
\node[blk, below right=10mm and -2mm of root] (sp3) {SP3 inductive\\ transfer (C2)};
\node[blkb, below=34mm of root] (sp4) {SP4 proactive (C4)};
\foreach \b in {sp1,sp2,sp3}\draw[flowarr] (root) -- (\b);
\draw[flowarr] (sp2) -- (sp4);
\node[font=\scriptsize, right=1mm of sp2] {cost: C3};
\end{tikzpicture}}
\caption{Decomposition of the amortization problem into four sub-problems, each mapped
to a measured claim.}
\label{fig:decomp}
\end{figure}

\begin{figure}[t]
\centering
\resizebox{0.82\columnwidth}{!}{%
\begin{tikzpicture}[x=8mm,y=8mm]
\foreach \i in {0,1,2,3}\foreach \j in {0,1,2,3}{
  \node[sat] (n\i\j) at (\i,\j) {};
}
\foreach \i in {0,1,2,3}\foreach \j in {0,1,2}{
  \draw[blue!55] (n\i\j) -- (n\i\the\numexpr\j+1\relax);
}
\foreach \i in {0,1,2}\foreach \j in {0,1,2,3}{
  \draw[black!45] (n\i\j) -- (n\the\numexpr\i+1\relax\j);
}
\begin{scope}[on background layer]
  \fill[orange!22] (1.6,-0.4) rectangle (2.4,3.4);
\end{scope}
\draw[flowarr, orange!80!black] (2.5,3.6) -- (3.1,3.6) node[right,font=\scriptsize]{drift $\Delta$};
\node[font=\scriptsize] at (2.0,-0.8) {hotspot band $\phi_t$};
\node[font=\scriptsize, blue!55!black] at (-0.9,1.5) {intra-plane};
\node[font=\scriptsize, black!55] at (1.5,-1.2) {cross-plane};
\end{tikzpicture}}
\caption{LEO $+$Grid ISL topology (torus). Intra-plane links (vertical) are stable;
cross-plane links (horizontal) vary with the geometry. The shaded band is the traffic
hotspot region, which drifts by a known $\Delta$ per slot.}
\label{fig:topo}
\end{figure}

\section{Method}\label{sec:method}
\subsection{Observable features (input to SP1)}
For a slot we form per-node features
$\phi_v=[\,\mathrm{out}_v,\,\mathrm{in}_v,\,x^{0,\mathrm{out}}_v,\,x^{0,\mathrm{in}}_v\,]$,
where $\mathrm{out}_v,\mathrm{in}_v$ are the rate originating and terminating at $v$, and
$x^{0,\cdot}_v$ aggregate the \emph{blind load} $x^{0}=\AoN(\prop)$ leaving and entering
$v$. Per-edge features are $[\,\prop_{uv},\,x^{0}_{uv}\,]$. The blind load is one $\AoN$
pass (no iteration) and is the most informative input, since the equilibrium price of a
link is largely a function of how overloaded that link would be under naive routing
(Section~\ref{sec:ablate-feat} quantifies this). All features are scaled to be size
invariant, so the model does not see absolute coordinates or counts and cannot shortcut
on them.

\subsection{Edge-price GNN (model for SP1)}
Node embeddings are initialized $h_v^{(0)}=\mathrm{ReLU}(W_{\mathrm{in}}\phi_v)$ and
updated for $l=0,\dots,K-1$ by a symmetric-normalized graph convolution
\begin{equation}
h_v^{(l+1)}=\mathrm{ReLU}\!\Big(W^{(l)}\!\!\sum_{u}\hat A_{vu}\,h_u^{(l)}\Big),
\label{eq:gcn}
\end{equation}
where $\hat A=\tilde D^{-1/2}(A{+}I)\tilde D^{-1/2}$ and $\tilde D$ is the degree of
$A{+}I$. A per-edge readout produces the price
\begin{equation}
\hat g_{uv}=\softplus\!\Big(\mathrm{MLP}\big([\,h_u,\,h_v,\,\prop_{uv},\,x^{0}_{uv}\,]\big)\Big).
\label{eq:readout}
\end{equation}
We use $K=3$ layers and hidden width $d=32$ throughout; the model has on the order of a
few thousand parameters, independent of $S$.

\subsection{Training objective}
We regress the log-price over a dataset $\mathcal{I}$ of (shell, demand) instances:
\begin{equation}
\mathcal{L}(\theta)=\frac{1}{|\mathcal{I}|}\sum_{i\in\mathcal{I}}\frac{1}{|E_i|}
\sum_{(u,v)\in E_i}\!\Big(\log(1{+}\hat g_{uv})-\log(1{+}g^{*}_{uv})\Big)^2 .
\label{eq:loss}
\end{equation}
The $\log(1{+}\cdot)$ scale tames the heavy tail of $g^{*}$ at saturated links, where the
BPR exponent makes the raw price span several orders of magnitude. Targets $g^{*}$ come
from solving UE \eqref{eq:msa} on the training instances offline; the cost of this
offline supervision is paid once and amortized over every deployment slot
(Algorithm~\ref{alg:train}).

\begin{algorithm}[t]
\caption{Offline supervision}
\label{alg:train}
\begin{algorithmic}[1]
\Require training shells, demand sampler, epochs $N$
\State $\mathcal{I}\gets\emptyset$
\For{each (shell, demand) instance}
  \State $x^{0}\gets\AoN(\prop)$;\ \ $\phi\gets\textsc{Features}(\mathcal{D},x^{0})$
  \State $x^{\mathrm{UE}}\gets$ MSA \eqref{eq:msa};\ \ $g^{*}\gets \cost(x^{\mathrm{UE}})/\prop-1$
  \State add $(\phi,g^{*})$ to $\mathcal{I}$
\EndFor
\For{$N$ epochs}
  \State update $\theta$ by gradient descent on $\mathcal{L}(\theta)$ \eqref{eq:loss}
\EndFor
\State \Return $\theta$
\end{algorithmic}
\end{algorithm}

\subsection{Supervision quality and the number of MSA iterations}
The quality of the learned price field is bounded by the quality of the targets $g^{*}$, so
the MSA solve used for supervision must converge well enough that the targets are close to the
true equilibrium. We use $T{=}20$ iterations of \eqref{eq:msa}, after which the average
relative change in the load is small and, as a downstream check, the supervised SO and UE
respect the price-of-anarchy ordering of Section~\ref{sec:pitfalls}. A larger $T$ tightens the
targets at higher offline cost but does not change the deployed model, whose inference is one
forward pass regardless; the iteration count is therefore a training-time knob, traded off once
and amortized over every slot. Because the offline solve is run on the small training shells,
not the large deployment shell, its cost is modest even though MSA itself does not scale, which
is the asymmetry the whole method exploits: pay for the equilibrium once, on a small instance,
and predict it everywhere else.

\subsection{Decoding and inference (SP2)}
Given $\hat g$, the routing is a single shortest-path assignment
$\pi=\AoN\big(\prop(1{+}\hat g)\big)$, shared across the whole demand matrix.
Algorithm~\ref{alg:infer} states one-slot inference: one blind pass to form the
features, one GNN forward, one decode pass.

\begin{algorithm}[t]
\caption{One-slot congestion-aware routing}
\label{alg:infer}
\begin{algorithmic}[1]
\Require graph $G_t$, demand $\mathcal{D}$ (or forecast $\hat{\mathcal{D}}$), weights $\theta$
\State $x^{0}\gets \AoN(\prop)$ \Comment{one blind pass = features}
\State $\phi\gets \textsc{Features}(\mathcal{D},x^{0})$
\State $\hat g\gets f_\theta(\phi,G_t)$ \Comment{one GNN forward, Eq.~\eqref{eq:readout}}
\State $\pi\gets \AoN\big(\prop(1{+}\hat g)\big)$ \Comment{one decode pass}
\State \Return $\pi$
\end{algorithmic}
\end{algorithm}

\subsection{Inductive transfer (SP3)}\label{sec:induct}
The parameters $\theta=\{W_{\mathrm{in}},W^{(l)},\mathrm{MLP}\}$ have dimensions set by
the feature width and hidden size only, not by $S$ or $|E|$. The aggregation
\eqref{eq:gcn} and readout \eqref{eq:readout} are defined pointwise over nodes and
edges. Hence $f_\theta$ evaluates on any graph, and a model trained on small shells
applies to a larger shell with no change, which is exactly SP3; Section~\ref{sec:c2}
measures the transfer to a $12\times$ larger shell.

\subsection{Proactive variant (SP4)}
Under known drift $\Delta$, the next-slot hotspot band is $\phi_t=\phi_{t-1}+\Delta$, so
the demand forecast $\hat{\mathcal{D}}_t$ is available. The proactive router decodes on
$f_\theta(\mathcal{S}(\hat{\mathcal{D}}_t))$; the reactive baseline decodes on
$f_\theta(\mathcal{S}(\mathcal{D}_{t-1}))$ and is therefore one step behind the moving
hotspot.

\subsection{Complexity (C3)}\label{sec:complexity}
A forward pass costs $O\!\big(K|E|d+K|V|d^2+|E|d^2\big)$ for hidden width $d$; a decode
pass costs $O\!\big(|\mathcal{D}|(|E|+|V|\log|V|)\big)$. The full router is one feature
pass, one forward, and one decode, i.e. two $\AoN$ passes plus the forward. UE by MSA
\eqref{eq:msa} costs $T$ $\AoN$ passes, $T\!\approx\!20$ in our runs. The predicted
speedup is thus $\Theta(T)$ up to the forward, and it \emph{grows} with scale because
the forward is cheap relative to the repeated routing on a larger graph; the measured
factor rises from $19\times$ at $132$ satellites to $29\times$ at $1584$
(Section~\ref{sec:c2}). Table~\ref{tab:cost} expresses the per-slot cost of each policy
in $\AoN$-pass units, the quantity that dominates at scale. Blind and geographic are one
pass; one-step is two; the GNN is two passes plus a forward that is asymptotically
dominated by a pass; UE and SO are $T$ passes. The GNN thus sits one additive pass above
the cheapest heuristics while delivering near-equilibrium quality, and an order of
magnitude below the equilibrium solve.

\begin{table}[t]
\centering
\caption{Per-slot routing cost in all-or-nothing ($\AoN$) shortest-path passes ($T$ is
the MSA iteration count, $\approx 20$). The GNN adds one forward to two passes; UE/SO
need $T$ passes.}
\label{tab:cost}
\begin{tabular}{lc}
\toprule
policy & cost (\AoN{} passes) \\
\midrule
blind, geographic & $1$ \\
one-step correction & $2$ \\
\textbf{GNN (ours)} & $2 + \text{1 forward}$ \\
UE, SO (MSA) & $T\approx 20$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Design rationale: why a supervised price field}\label{sec:rationale}
Three design choices distinguish this approach from the alternatives. First, we predict
a \emph{price field} rather than a routing directly: a single $|E|$-dimensional field is
shared by every demand and decoded by an existing, trusted shortest-path router, so the
output is always a valid routing and inductivity across constellation size is immediate.
Predicting per-flow paths, by contrast, would scale with the demand matrix and require a
learned decoder. Second, we \emph{supervise} on the equilibrium rather than learn by
reinforcement: the UE target is computable offline, the objective \eqref{eq:loss} is a
stable regression, and there is no exploration cost or reward-shaping at deployment,
which matters when each slot lasts seconds. Third, we predict the equilibrium of a
\emph{convex} traffic-assignment problem, so the target is unique and well posed; the
network learns to amortize an optimization it could otherwise have to solve from scratch
each slot.

\section{Pitfalls in Evaluating Learned LEO Traffic Engineering}\label{sec:pitfalls}
Before reporting results we document three evaluation pitfalls we encountered. Each can
silently fabricate or distort a headline number; each has a simple fix. We state them
because they are easy to commit in this setting and because our corrected estimator is
what produces the numbers in Section~\ref{sec:results}.

\textbf{P1: survivorship in greedy baselines.} A natural baseline, greedy geographic
routing (forward to the neighbor closest to the destination), stalls in local minima on
the $+$Grid ISL graph and fails to deliver $53$ to $68\%$ of flows in our loaded
regime. Scoring only the delivered flows credits the baseline with a near-optimal mean
delay, because the survivors are the short, easy paths. \emph{Fix:} guarantee delivery
(here, a shortest-path fallback for stalled flows) and report a survivorship-free total
travel time over all demands.

\textbf{P2: scaling claims on a moving denominator.} An early claim that an operator's
advantage ``grows with the constellation'' rested on a relative-percentage gain whose
denominator (a baseline error) itself grew super-linearly, so a shrinking percentage
masked a widening absolute gap and vice versa. \emph{Fix:} judge scaling on the
decision-relevant absolute quantity, not on a ratio with a moving denominator.

\textbf{P3: measuring an equilibrium off its operating point.} The travel time of UE or
SO must be evaluated on the converged (MSA-averaged) load using the edge form of
\eqref{eq:ttt}. Re-routing a single all-or-nothing pass on the converged cost and
scoring that instead concentrates flow off the equilibrium split and inflates the
reference: in our data it overstated UE travel time by up to $5\times$ (a relative
travel time of $0.23$ versus the correct $0.05$ at high load), which would have made the
learned router look closer to optimal than it is. \emph{Fix:} evaluate
$\TTT=\sum_{(u,v)} x_{uv}\cost_{uv}(x_{uv})$ on the converged load. As a check, the
corrected estimator respects $\TTT(\mathrm{SO})\le\TTT(\mathrm{UE})$ at every load, with
the price of anarchy growing from $1.00$ to $1.11$ as load rises; the inflated estimator
violated this ordering, which is how we caught it.

None of these three pitfalls is specific to LEO or to graph networks. P1 arises whenever a
baseline can decline to act and the metric ignores the declined cases; P2 arises whenever a
claim about a trend is read off a normalized quantity with a shifting denominator; P3 arises
whenever an iterative solver is scored on a re-derived rather than its converged state. The
general lesson is to fix a single survivorship-free, absolute, operating-point-correct figure
of merit before comparing methods, and to keep a cheap internal consistency check (here, the
price-of-anarchy ordering) that a corrupted estimator will fail. We surface these as a
reusable evaluation protocol for learned traffic engineering, since each silently produced a
publishable-looking but wrong number before it was caught.

\section{Experimental Setup}\label{sec:setup}
We simulate Walker shells with a pure-kinematic propagator (no packet-level dependency,
fully reproducible from parameters): an Iridium-like shell ($66$ satellites, diameter
$9$) and Starlink-like $53^\circ$ shells of $132$ (diameter $16$), $264$ ($28$), and
$1584$ satellites. Capacity is uniform; demands follow the drifting gravity model. We
report total travel time \eqref{eq:ttt} relative to blind shortest path (lower is
better; blind $=1.00$), averaged over multiple demand instances and seeds with standard
deviations. Baselines: \emph{blind}; \emph{geographic} greedy (great-circle next hop,
with the P1 fallback); \emph{one-step} congestion correction (route on one blind-load
pass); \emph{GNN} (ours); and the references \emph{UE} and \emph{SO}. Learned results are
param-matched across operators and reported with error bars; all metrics are
survivorship free per Section~\ref{sec:pitfalls}. The GNN is trained on the $132$-shell
and evaluated both in distribution and zero shot on larger shells.

\section{Results}\label{sec:results}

\subsection{Headroom for congestion awareness}\label{sec:headroom}
Figure~\ref{fig:headroom} sweeps offered load. At light load, blind shortest path is
near optimal and UE offers nothing. As load grows, blind concentrates traffic on
bottleneck links (its peak utilization reaches up to nine times capacity while
alternates idle) and the gain of UE over blind rises to $9$, $55$, $90$, and $98\%$.
The congestion task, not static routing, is where a learned router helps.

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[
  xlabel={offered load (number of demands)}, ylabel={UE travel-time gain (\%)},
  xmin=0,xmax=1700, ymin=0, ymax=100, grid=both, legend pos=south east,
  width=8.4cm, height=5.4cm, tick align=outside]
\addplot[mark=*,thick,blue] coordinates {(100,0.1)(300,9.4)(600,55.4)(1000,89.9)(1600,97.6)};
\addlegendentry{$132$-sat shell}
\addplot[mark=square*,thick,red,dashed] coordinates {(100,0.0)(300,10.5)(600,61.9)(1000,92.0)(1600,97.9)};
\addlegendentry{$264$-sat shell}
\end{axis}
\end{tikzpicture}}
\caption{Congestion headroom: the gain of the user equilibrium over congestion-blind
shortest path grows with load, to $98\%$ under heavy load. None at light load.}
\label{fig:headroom}
\end{figure}

Table~\ref{tab:headroom} gives the absolute numbers behind the figure for the
$132$-shell. The ratio $\TTT(\mathrm{blind})/\TTT(\mathrm{UE})$ widens from $1.0$ at
light load to about $41\times$ at the heaviest, because the BPR exponent $\beta{=}4$
makes the cost on an overloaded link grow as the fourth power of its utilization, which
blind routing drives to nine times capacity on the bottleneck while leaving parallel
links idle. Crucially, this is a load-balancing effect, not a global capacity shortfall:
the same demand carried over diverse paths is comfortably feasible, which is why a
congestion-aware policy recovers almost all of the gap.

\begin{table}[t]
\centering
\caption{Absolute travel time (arbitrary units) of blind versus UE on the $132$-shell as
offered load grows, and their ratio. Mean over seeds.}
\label{tab:headroom}
\begin{tabular}{lccccc}
\toprule
demands & $100$ & $300$ & $600$ & $1000$ & $1600$ \\
\midrule
blind & $4.8$ & $17.2$ & $78.6$ & $796.6$ & $7433.5$ \\
UE    & $4.8$ & $15.5$ & $35.0$ & $79.7$  & $180.6$ \\
ratio & $1.0$ & $1.1$  & $2.2$  & $10.0$  & $41.2$ \\
\bottomrule
\end{tabular}
\end{table}

The corrected estimator of Section~\ref{sec:pitfalls} also lets us read off the price of
anarchy. Solving SO by the marginal-cost iteration and comparing to UE,
Table~\ref{tab:poa} shows $\TTT(\mathrm{SO})\le\TTT(\mathrm{UE})$ at every load, with the
ratio rising from $1.00$ at light load to $1.11$ under heavy load. The two operating
points nearly coincide, so the learned router's target (UE) is within a few percent of
the true system optimum, and matching UE is matching the optimum for practical purposes.

\begin{table}[t]
\centering
\caption{Price of anarchy $\TTT(\mathrm{UE})/\TTT(\mathrm{SO})$ on the $132$-shell;
$\ge 1$ at every load confirms the corrected estimator (Section~\ref{sec:pitfalls}).}
\label{tab:poa}
\begin{tabular}{lccc}
\toprule
demands & $300$ & $600$ & $1000$ \\
\midrule
$\TTT(\mathrm{UE})/\TTT(\mathrm{SO})$ & $1.00$ & $1.03$ & $1.11$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The GNN approaches UE and beats every cheap baseline (C1)}
Figure~\ref{fig:baselines} and Table~\ref{tab:baselines} report TTT relative to blind. In
distribution, the GNN ($0.43$) closely approaches UE ($0.38$), recovering about $92\%$ of
the blind-to-UE gain; SO ($0.37$) and UE nearly coincide, so the target is essentially
the system optimum. The cheap alternatives fail: one-step correction overshoots and
oscillates (relative travel time $1.07$, \emph{worse} than blind, because rerouting onto
the apparently idle alternates creates fresh hotspots), and geographic greedy stalls on
the ISL grid, needing the fallback for $55\%$ of flows and still reaching only $0.79$. The contrast
between the GNN and one-step is the central evidence that the problem is not trivial: a
single congestion-blind pass tells you where the hotspots are, but reacting to it by
rerouting everyone onto the apparent alternates simply relocates the hotspots, a
one-step oscillation that overshoots the equilibrium. The GNN, trained on the converged
prices, predicts the \emph{fixed point} of that process directly and so avoids the
overshoot. This is why a learned price field, rather than a hand-tuned correction, is
needed.

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[
  ybar, bar width=6pt, width=8.6cm, height=5.6cm,
  symbolic x coords={blind,geographic,one-step,GNN,UE,SO},
  xtick=data, x tick label style={font=\small,rotate=20,anchor=east},
  ylabel={TTT relative to blind}, ymin=0, ymax=1.2, grid=both,
  legend pos=north east, enlarge x limits=0.12]
\addplot[fill=blue!30] coordinates {(blind,1.00)(geographic,0.79)(one-step,1.07)(GNN,0.43)(UE,0.38)(SO,0.37)};
\addplot[fill=orange!50] coordinates {(blind,1.00)(geographic,0.79)(one-step,0.54)(GNN,0.37)(UE,0.05)(SO,0.05)};
\legend{in-distribution, OOD ($264$)}
\end{axis}
\end{tikzpicture}}
\caption{Travel time relative to blind (lower is better). The GNN approaches UE in
distribution and is the best \emph{deployable} policy out of distribution; the cheap
heuristics fail (one-step is worse than blind in distribution, geographic stalls). UE and
SO nearly coincide.}
\label{fig:baselines}
\end{figure}

\begin{table}[t]
\centering
\caption{Total travel time relative to blind (lower is better; blind $=1.00$). Mean over
instances and seeds. SO/UE are references; the GNN is the deployable policy.}
\label{tab:baselines}
\begin{tabular}{lccccc}
\toprule
split & SO & UE & geographic & one-step & \textbf{GNN} \\
\midrule
in-distribution & $0.37$ & $0.38$ & $0.79$ & $1.07$ & $\mathbf{0.43}$ \\
OOD ($264$) & $0.05$ & $0.05$ & $0.79$ & $0.54$ & $\mathbf{0.37}$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Inductive transfer to a $12\times$ larger shell, and cost (C2, C3)}\label{sec:c2}
Trained only on the $132$-satellite shell, the model is applied zero shot to the
$264$- and $1584$-satellite shells. Table~\ref{tab:scale} reports the recovered fraction
of the blind-to-UE gain and the inference speedup over the UE solve. The GNN recovers
$65\%$ at $264$ and $71\%$ at $1584$ (a $12\times$ size jump), remaining the best
deployable policy throughout, while the speedup over MSA \emph{rises} from $19\times$ to
$29\times$ (Fig.~\ref{fig:speedup}). The amortization therefore pays off more, not less,
as the constellation grows, which is the regime that matters operationally. Two points
deserve emphasis. First, the transfer is genuinely zero shot: the $1584$-shell is never
seen during training, has a different number of planes and satellites per plane, and a
larger diameter, yet the same weights produce a near-equilibrium price field on it. This
works because the learned map is from local load structure to a local price, a relation
that is scale invariant in a way the absolute routing is not. Second, the recovered
fraction at $1584$ ($71\%$) is in fact higher than at $264$ ($65\%$); we attribute this
to the lighter evaluation configuration used at the largest scale rather than to a true
non-monotonicity, and we report both with their standard deviations rather than select
the more favorable one. In all cases the GNN remains strictly the best deployable policy,
which is the operational claim.

\begin{table}[t]
\centering
\caption{Inductive transfer (train on $132$) and inference cost. Recovered fraction of
the blind-to-UE gain (mean $\pm$ std), and one-shot GNN routing versus the MSA solve.}
\label{tab:scale}
\begin{tabular}{lcccc}
\toprule
shell & sats & recovered \% & GNN (s) & speedup vs UE \\
\midrule
in-dist & $132$ & $88\pm4$ & $0.05$ & $19\times$ \\
OOD & $264$ & $65\pm9$ & $0.25$ & $20\times$ \\
OOD & $1584$ & $71\pm8$ & $8.4$ & $29\times$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\resizebox{0.86\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[
  xlabel={shell size (satellites)}, ylabel={speedup of GNN over UE/MSA ($\times$)},
  xmin=0,xmax=1700, ymin=0, ymax=34, grid=both, width=8.2cm, height=5.0cm,
  tick align=outside]
\addplot[mark=*,thick,blue] coordinates {(132,18.6)(264,20.4)(1584,29.0)};
\end{axis}
\end{tikzpicture}}
\caption{Inference speedup of the one-shot GNN router over the MSA equilibrium solve,
versus shell size. The advantage grows with the constellation.}
\label{fig:speedup}
\end{figure}

\subsection{Proactive routing under drift (C4)}
Figure~\ref{fig:proactive} sweeps the per-slot hotspot drift. At zero drift the reactive
and proactive routers are identical, as they must be. As the hotspot moves, the reactive
lag-1 router routes for yesterday's load and degrades, passing blind at moderate drift
and reaching $1.6\times$ blind at the largest drift. The proactive router, fed the
predicted demand, tracks the clairvoyant equilibrium throughout; the proactive advantage
grows from $0$ to $76\%$. The zero-drift control is important: it confirms that the
proactive gain is genuinely due to anticipating the hotspot and not to any artifact of
the two pipelines, which are identical when there is nothing to anticipate. The crossover
where the reactive router passes blind (around a per-slot drift of one radian in our
sweep) marks the operating point beyond which a stale congestion estimate is worse than
ignoring congestion altogether, a concrete criterion for when proactivity stops being
optional. Because the GNN's input is just a demand-derived feature, switching from
reactive to proactive requires no change to the model, only feeding it the forecast
instead of the last slot.

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[
  xlabel={per-slot hotspot drift (rad)}, ylabel={TTT relative to blind},
  xmin=0,xmax=1.55, ymin=0.3, ymax=1.75, grid=both, legend pos=north west,
  width=8.4cm, height=5.6cm, tick align=outside]
\addplot[mark=*,thick,red] coordinates {(0,0.50)(0.3,0.59)(0.6,0.77)(1.0,1.02)(1.5,1.63)};
\addlegendentry{reactive (lag-1)}
\addplot[mark=square*,thick,blue] coordinates {(0,0.50)(0.3,0.52)(0.6,0.54)(1.0,0.44)(1.5,0.38)};
\addlegendentry{proactive (predicted)}
\addplot[mark=triangle*,thick,black,densely dashed] coordinates {(0,0.44)(0.3,0.44)(0.6,0.44)(1.0,0.39)(1.5,0.35)};
\addlegendentry{UE (clairvoyant)}
\addplot[dotted,very thick,gray] coordinates {(0,1.0)(1.5,1.0)};
\addlegendentry{blind}
\end{axis}
\end{tikzpicture}}
\caption{Proactive versus reactive routing under drifting hotspots. The reactive lag-1
router degrades past blind as the hotspot moves; the proactive router tracks the
clairvoyant optimum.}
\label{fig:proactive}
\end{figure}

\subsection{Operating regime}\label{sec:regime}
The value of the method is regime dependent, and Table~\ref{tab:headroom} makes the boundary
explicit. Below an average utilization of about one, blind shortest path is within a few
percent of optimal and there is nothing to amortize; a deployment that is always lightly loaded
does not need this machinery. The learned router earns its place in the loaded regime, where the
blind-to-UE gap opens to tens of percent and beyond, and where recomputing the equilibrium each
slot would otherwise be the cost driver. This is also the regime a capacity-planned network
spends its busy hours in, and the regime a moving hotspot creates locally even when the network
as a whole is moderately loaded, since the hotspot concentrates demand onto a small cut. The
single trained model spans the whole range: at light load it predicts a near-uniform,
near-zero price field and reduces to blind routing, and as load rises it predicts the
increasingly structured prices that spread traffic off the bottlenecks, with no regime switch
or retuning.

\subsection{Feature ablation}\label{sec:ablate-feat}
The blind-load feature is what makes the price predictable. With only per-node demand
intensity as input, the model must infer where congestion will concentrate purely from the
demand pattern and the graph structure, a global and ambiguous inference; in that
configuration the zero-shot recovered fraction falls to roughly half of the reported value
and its variance rises several fold, and some seeds route worse than blind. Supplying the
blind load, the loads that one congestion-blind pass would induce, turns the task into a
largely local correction: a link that the naive pass would overload is exactly the link that
needs a higher price, and the GNN's neighborhood aggregation refines this with the structure
around it. The feature costs a single $\AoN$ pass to compute, the same pass already used to
build the node features, so it is effectively free. This also explains why the global
propagation operators of Section~\ref{sec:ablation} do not help once the feature is present:
the long-range information they were meant to gather is already summarized, locally, by the
blind load.

\subsection{Operator ablation, including the quantum walk (negative)}\label{sec:ablation}
Table~\ref{tab:ablation} compares three param-matched propagation operators: a local
graph convolution (GCN), a global classical-diffusion operator (Heat), and a global
quantum-walk operator (QW). Once the blind-load features are present, predicting the
price is largely a local function of a link's own load, so the plain GCN is best out of
distribution and the global operators add variance without benefit; the quantum-walk
operator is the most variable ($\pm 20$). The quantum walk originally motivated this work
as a way to capture long-range structure; the data did not support it, and we report it
as an honest negative ablation rather than force it into the contribution.

\begin{table}[t]
\centering
\caption{Operator ablation: recovered fraction of the blind-to-UE gain (mean $\pm$ std,
$n{=}12$). The plain GCN is best out of distribution; the global quantum-walk operator
does not help and is the most variable.}
\label{tab:ablation}
\begin{tabular}{lcc}
\toprule
operator & in-distribution & OOD ($264$) \\
\midrule
\textbf{GCN (local)} & $86\pm6$ & $\mathbf{65\pm9}$ \\
Heat (global diffusion) & $89\pm4$ & $52\pm6$ \\
QW (global quantum walk) & $82\pm8$ & $62\pm20$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Deployment Considerations}
In an operational control loop the router runs once per slot on the known next-slot
topology. The only online inputs are the demand estimate and one congestion-blind
shortest-path pass to form features; the GNN forward and the decode pass follow. Because
the topology is known ahead of time, the feature pass and the forward can be precomputed
for the upcoming slot, leaving only the decode on the live demand. The model is small (a
few thousand parameters) and its weights are fixed across constellation sizes, so a
single trained model serves an entire program of shells. Two safeguards follow naturally
from our analysis. When the demand forecast is unreliable, for example during an
unmodeled surge, the controller can fall back to reactive features or to blind routing,
which our proactive sweep shows is preferable to acting on a badly stale estimate beyond
the crossover drift. And because the decoded routing is always a valid shortest-path
assignment on a positive cost, it can never produce a loop or a disconnected route,
unlike a directly predicted next-hop policy that must be guarded against such failures.

\section{Threats to Validity}
We separate threats we controlled from those that remain. \emph{Construct validity}:
total travel time under the BPR cost is the standard traffic-assignment objective and is
survivorship free by Section~\ref{sec:pitfalls}, but it is a flow-level proxy for the
packet-level delay an operator ultimately cares about. \emph{Internal validity}: every
learned number is param-matched across operators, averaged over multiple instances and
seeds with standard deviations, and features are scale-normalized so the model cannot
shortcut on absolute coordinates; the operator ablation isolates the propagation operator
at a fixed parameter budget. \emph{External validity}: results are on synthetic Walker
shells with a uniform capacity and a gravity demand model, and the zero-shot transfer was
tested up to a $12\times$ size jump and one shell geometry; other geometries,
heterogeneous link capacities, and real traffic traces remain to be studied.
\emph{Reproducibility}: the pipeline is deterministic from parameters and every figure
number is regenerated from a result CSV by a released script.

\section{Discussion and Limitations}
The simulator is kinematic and flow level, not packet level; it captures
geometry-driven delay and load-induced congestion through the BPR model, the mechanism
the method targets, but it does not model queueing dynamics or protocol behavior, and a
packet-level study (for example on Hypatia) is the natural next validation. UE is a
selfish equilibrium and only a lower bound away from SO; in our regime the two nearly
coincide (price of anarchy up to $1.11$), so the learned target is near-optimal, but at
other capacity-to-demand ratios the gap could matter and SO would be the better target.
The zero-shot transfer leaves a residual gap to UE on the larger shell; training across
a range of shell sizes, rather than a single small shell, is a natural way to close it.
The proactive variant assumes the demand drift is predictable, which holds for the
diurnal rotation of the dense-traffic band but not for sudden, unforecastable surges,
where a reactive fallback would be needed. Finally, our routing decode is a single
shortest-path assignment; splitting flow across multiple paths, as UE does, could narrow
the remaining gap at the cost of a more complex decoder.

\section{Future Work}
Several directions follow directly. A packet-level study, for example on
Hypatia~\cite{hypatia}, would replace the flow-level BPR proxy with queueing and protocol
dynamics and test whether the predicted price field still routes well when delay is
realized by buffers rather than a cost function. A multipath decoder, splitting each demand
across the few shortest paths under the predicted cost as the equilibrium itself does, is the
most direct route to closing the residual gap to UE, at the cost of a more complex decode.
Heterogeneous and time-varying capacities, including feeder-link and ground-segment
bottlenecks, would stress the assumption of a uniform ISL capacity and likely make the price
field richer and more valuable to learn. Training across a spectrum of shell sizes, rather
than a single small shell, should improve the zero-shot transfer and reduce its variance. On
the proactive side, replacing the known-drift forecast with a learned demand predictor would
let the method handle hotspots whose motion is only statistically predictable, with the
reactive and blind fallbacks of Section~\ref{sec:rationale} as a safety net when the forecast
is poor. Finally, light online fine-tuning of the released weights on a target constellation
could recover the few points of optimality lost in zero-shot transfer while preserving the
one-shot inference cost.

\section{Conclusion}
We presented an inductive graph network that amortizes congestion-aware traffic
engineering for LEO mega-constellations. After formalizing routing under load as the
amortization of a traffic-assignment equilibrium and decomposing it into four
sub-problems, we solved the price-prediction core with a GNN that predicts the
user-equilibrium congestion price on every ISL in one forward pass from cheap load
features. It approaches the equilibrium in distribution, transfers zero shot to a
$12\times$ larger shell as the best deployable policy, runs up to $29\times$ faster than
the iterative solve with the saving growing at scale, and, fed predicted demand,
sustains low travel time under drifting hotspots where reactive control fails. We
documented three evaluation pitfalls and a protocol to avoid them, and we reported why
an initially appealing quantum-walk operator did not earn its place. The full pipeline,
including the negative results, is released so that the methodology is reusable.

\section*{Data and Code Availability}
The simulator, models, experiments, and figure-generation scripts are released. The
constellation and traffic models are in \texttt{sim/}; the edge-price GNN and the
experiments producing each claim are in \texttt{experiments/} (\texttt{p4} headroom,
\texttt{p5} router and operator ablation, \texttt{p6} baselines and timing, \texttt{p7}
proactive); every figure number is aggregated from \texttt{results/*.csv} by
\texttt{paper/make\_figs\_data.py}. All results are reproducible from parameters with no
external data dependency.

\appendices
\section{Reproducibility notes}
All learned results use $K{=}3$ layers, hidden width $32$, the Adam optimizer, and are
averaged over multiple demand instances and random seeds with standard deviations
reported. The UE/SO references are solved by MSA with $20$ iterations; the eigendecomposition
required only by the global operators is skipped for the GCN, which is what makes the
$1584$-satellite evaluation tractable. Features are scale-normalized so the model cannot
shortcut on absolute coordinates or counts.

\begin{thebibliography}{99}
\bibitem{handley2018} M. Handley, ``Delay is not an option: Low latency routing in
space,'' in \emph{Proc. ACM HotNets}, 2018.
\bibitem{bhattacherjee2019} D. Bhattacherjee and A. C. Snoeren, ``Network topology
design at 27,000 km/hour,'' in \emph{Proc. ACM CoNEXT}, 2019.
\bibitem{hypatia} S. Kassing, D. Bhattacherjee, A. B. Águas, J. E. Saethre, and
A. Singla, ``Exploring the `Internet from space' with Hypatia,'' in \emph{Proc. ACM
IMC}, 2020.
\bibitem{starperf} Z. Lai, H. Li, and J. Li, ``StarPerf: Characterizing network
performance for emerging mega-constellations,'' in \emph{Proc. IEEE ICNP}, 2020.
\bibitem{gcn} T. N. Kipf and M. Welling, ``Semi-supervised classification with graph
convolutional networks,'' in \emph{Proc. ICLR}, 2017.
\bibitem{gat} P. Veli\v{c}kovi\'c \emph{et al.}, ``Graph attention networks,'' in
\emph{Proc. ICLR}, 2018.
\bibitem{graphsage} W. L. Hamilton, R. Ying, and J. Leskovec, ``Inductive
representation learning on large graphs,'' in \emph{Proc. NeurIPS}, 2017.
\bibitem{routenet} K. Rusek \emph{et al.}, ``RouteNet: Leveraging graph neural networks
for network modeling and optimization in SDN,'' \emph{IEEE J. Sel. Areas Commun.}, vol.
38, no. 10, 2020.
\bibitem{almasan} P. Almasan \emph{et al.}, ``Deep reinforcement learning meets graph
neural networks: Exploring a routing optimization use case,'' \emph{Computer
Communications}, vol. 196, 2022.
\bibitem{wardrop} J. G. Wardrop, ``Some theoretical aspects of road traffic research,''
\emph{Proc. Institution of Civil Engineers}, 1952.
\bibitem{bpr} Bureau of Public Roads, \emph{Traffic Assignment Manual}, U.S. Dept. of
Commerce, 1964.
\bibitem{beckmann} M. Beckmann, C. B. McGuire, and C. B. Winsten, \emph{Studies in the
Economics of Transportation}, Yale Univ. Press, 1956.
\bibitem{poa} T. Roughgarden and \'E. Tardos, ``How bad is selfish routing?''
\emph{Journal of the ACM}, vol. 49, no. 2, 2002.
\bibitem{valadarsky} A. Valadarsky, M. Schapira, D. Shahaf, and A. Tamar, ``Learning to
route,'' in \emph{Proc. ACM HotNets}, 2017.
\bibitem{xu} Z. Xu \emph{et al.}, ``Experience-driven networking: A deep reinforcement
learning based approach,'' in \emph{Proc. IEEE INFOCOM}, 2018.
\bibitem{qwnn} S. Dernbach, A. Mohseni-Kabir, S. Pal, and D. Towsley, ``Quantum walk
neural networks,'' in \emph{Complex Networks}, 2018.
\end{thebibliography}

\vspace{-2mm}
\begin{IEEEbiographynophoto}{Do Phuc Hao}
received the Ph.D. degree in telecommunications from a university in Russia. He is with
the Center for Artificial Intelligence Research and Application (CAIRA), Da Nang
Architecture University, Da Nang, Vietnam. His research interests include
non-terrestrial networks, graph learning for networking, and trustworthy AI.
\end{IEEEbiographynophoto}

\end{document}
